---
title: "Cyclistic"
author: "Konrad Peschka"
date: "04/02/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#--------------------------------------------
#-------------------------------------------
#    1- Prepare libraries and Data Frame
#-------------------------------------------
#--------------------------------------------

#---------------------------------
#   1.A - Install needed packages
#--------------------------------
```{r}
  install.packages("tidyverse")
  install.packages("lubridate")
  install.packages("geosphere")
  install.packages("ggplot2")

  library(tidyverse)
  library(lubridate)
  library(geosphere)
  library(ggplot2)
```
#----------------------------------------
#    1.B- Import and combine the Data
#----------------------------------------

```{r}
#Import data
  trip_data_2022_01 <- read.csv("2022-01-divvy-tripdata.csv")
  trip_data_2022_02 <- read.csv("2022-02-divvy-tripdata.csv")
  trip_data_2022_03 <- read.csv("2022-03-divvy-tripdata.csv")
  trip_data_2022_04 <- read.csv("2022-04-divvy-tripdata.csv")
  trip_data_2022_05 <- read.csv("2022-05-divvy-tripdata.csv")
  trip_data_2022_06 <- read.csv("2022-06-divvy-tripdata.csv")
  trip_data_2022_07 <- read.csv("2022-07-divvy-tripdata.csv")
  trip_data_2022_08 <- read.csv("2022-08-divvy-tripdata.csv")
  trip_data_2022_09 <- read.csv("2022-09-divvy-tripdata.csv")
  trip_data_2022_10 <- read.csv("2022-10-divvy-tripdata.csv")
  trip_data_2022_11 <- read.csv("2022-11-divvy-tripdata.csv")
  trip_data_2022_12 <- read.csv("2022-12-divvy-tripdata.csv")
```

#Print all the columns to check it visually 
```{r}
  colname_01 <- colnames(trip_data_2022_01)
  colname_02 <- colnames(trip_data_2022_02)
  colname_03 <- colnames(trip_data_2022_03)
  colname_04 <- colnames(trip_data_2022_04)
  colname_05 <- colnames(trip_data_2022_05)
  colname_06 <- colnames(trip_data_2022_06)
  colname_07 <- colnames(trip_data_2022_07)
  colname_08 <- colnames(trip_data_2022_08)
  colname_09 <- colnames(trip_data_2022_09)
  colname_10 <- colnames(trip_data_2022_10)
  colname_11 <- colnames(trip_data_2022_11)
  colname_12 <- colnames(trip_data_2022_12)
```

  #if you wanna test the code modify this colname_2 in some colnames
```{r}
  #colname_02 <- c("ride_id","rideable_type2","started_at","ended_at2","start_station_name","start_station_id","end_station_name","end_station_id","start_lat","start_lng","end_lat","end_lng","member_casual")
  ```

#Run a code to check column names to ensure we can join all the data

```{r}
length_of_colnames <- length(colname_01) #With this variable i can know the leght of the vector
i=1 #it allways start in 1

#here you need to select mannually each column to be comparated with colname_1

colname_b <- colname_02
n=2 #the n comes with the colname_2
#colname_b <- colname_03
#n=3
#colname_b <- colname_04
#n=4
#colname_b <- colname_05
#n=5
#colname_b <- colname_06
#n=6
#colname_b <- colname_07
#n=7
#colname_b <- colname_08
#n=8
#colname_b <- colname_09
#n=9
#colname_b <- colname_10
#n=10
#colname_b <- colname_11
#n=11
#colname_b <- colname_12
#n=12

print(paste0("Comparing: colname_01 = colname_",n))
while(i<(length_of_colnames+1)) #check columns by columns if they are the same
{	
	if(colname_01[i] == colname_b[i]){
		print(paste0("Column ", i," - Equal"))
	}
	else
	{
		print(paste0("Column ", i," - NotEqual"))
	}
	i <- i+1
}
```

#Combine data into one set.
```{r}
all_data<- rbind(trip_data_2022_01, trip_data_2022_02, trip_data_2022_03, trip_data_2022_04, trip_data_2022_05, trip_data_2022_06, trip_data_2022_07, trip_data_2022_08, trip_data_2022_09, trip_data_2022_10, trip_data_2022_11, trip_data_2022_12)
```

#----------------------------------------
#    1.C- Cast and Clean Data Frame
#----------------------------------------
#summary(all_data)

# Cast the data from CHAR to DATE
# Convert the "started_at" and "ended_at" columns drom char to Date and Time
```{r}
all_data$started_at <- strptime(all_data$started_at, "%Y-%m-%d %H:%M:%S")
all_data$ended_at <- strptime(all_data$ended_at, "%Y-%m-%d %H:%M:%S")

all_data <- all_data %>% drop_na()
```

#----------------------------------------------------
#----------------------------------------------------
#   2- Analyze Annual ans Casual member separatly
#----------------------------------------------------
#----------------------------------------------------

#-------------------------
#   2.A - Annual member
#-------------------------

# a- Create a dataframe for annual riders
```{r}
annual_members <- all_data[all_data$member_casual == "member", ]
```

# b- Calculate the total number of trips
```{r}
total_trips_annual <- nrow(annual_members)
```

# c- Calculate the average trip duration
```{r}
trip_duration_annual_sec <- as.numeric(mean(annual_members$ended_at - annual_members$started_at))
average_duration_annual_min <- trip_duration_annual_sec / 60
```

# d- Calculate distance
# Filter out the trips with missing latitude or longitude data
```{r}
valid_rows <- complete.cases(annual_members[, c("start_lng", "start_lat", "end_lng", "end_lat")])
valid_```

# Calculate the distances between the start and end points
```{r}
trip_distances_annual <- distHaversine(cbind(valid_annual_members$start_lng, valid_annual_members$start_lat), cbind(valid_annual_members$end_lng, valid_annual_members$end_lat))
```

# Calculate the average distance
```{r}
average_distance_annual_m <- mean(trip_distances_annual, na.rm = TRUE)
average_distance_annual_km <- average_distance_annual_m / 1000
```

# e- Print the results
```{r}
result_string_1 <- sprintf("The average trip duration is %0.1f minutes.", average_duration_annual_min)
result_string_2 <- sprintf("The average trip distance is %0.1f kilometers.", average_distance_annual_km)

print("Results for Annual Member")
print(paste("Total trips:", total_trips_annual))
print(result_string_1)
print(result_string_2)
```

#-------------------------
#   2.B - Casual member
#-------------------------

# Create a dataframe for casual riders
```{r}
casual_riders <- all_data[all_data$member_casual == "casual", ]

#Calculate the total number of trips
```{r}
total_trips_casual <- nrow(casual_riders)
```

# Calculate the average trip duration
```{r}
trip_duration_casual_sec <- as.numeric(mean(casual_riders$ended_at - casual_riders$started_at))
average_duration_casual_min <- trip_duration_casual_sec / 60
```

# Filter out the trips with missing latitude or longitude data
```{r}
valid_rows <- complete.cases(casual_riders[, c("start_lng", "start_lat", "end_lng", "end_lat")])
valid_casual_riders <- casual_riders[valid_rows, ]
```

# Calculate the distances between the start and end points
```{r}
trip_distances_casual <- distHaversine(cbind(valid_casual_riders$start_lng, valid_casual_riders$start_lat), cbind(valid_casual_riders$end_lng, valid_casual_riders$end_lat))
```

# Calculate the average distance
```{r}
average_distance_casual_m <- mean(trip_distances_casual, na.rm = TRUE)
average_distance_casual_km <- average_distance_casual_m / 1000
```

# Print the results
```{r}
result_string_3 <- sprintf("The average trip duration is %0.1f minutes.", average_duration_casual_min)
result_string_4 <- sprintf("The average trip distance is %0.1f kilometers.", average_distance_casual_km)

print("Results for Casual Member")
print(paste("Total trips:", total_trips_casual))
print(result_string_3)
print(result_string_4)
```

#------------------------------------------------
#------------------------------------------------
#   3- Analyze both type of members together
#------------------------------------------------
#------------------------------------------------

#--------------------------------------------
#   3.A - Member behavior vs day of week
#--------------------------------------------

# Bar chart showing the difference in behavior between casual riders and annual members for each day of theweek:
# Combine the data for casual riders and annual members into a single data frame
```{r}
day_of_week_rides <- rbind(
  casual_riders %>% 
    mutate(day_of_week = as.factor(weekdays(as.Date(started_at))), rider_type = "Casual Rider") %>% 
    group_by(day_of_week, rider_type) %>% 
    summarize(count = n()),
  annual_members %>% 
    mutate(day_of_week = as.factor(weekdays(as.Date(started_at))), rider_type = "Annual Member") %>% 
    group_by(day_of_week, rider_type) %>% 
    summarize(count = n())
)
```

# Plot the bar chart
```{r}
day_of_week_rides$day_of_week <- factor(day_of_week_rides$day_of_week, 
                                        levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

ggplot(data = day_of_week_rides, aes(x = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), y = count, fill = rider_type)) +
  geom_boxplot(outlier.size = 0) +
  labs(x = "Day of Week", y = "Number of Rides", title = "Number of Rides by Day of Week and Rider Type") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("red", "blue"), labels = c("Casual Riders", "Annual Members")) +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
```

#------------------------------------------------
#   3.B - Member behavior vs month of the year
#------------------------------------------------

#Difference in behavior between casual riders and annual members for each month of the year:
# Combine the data for casual riders and annual members into a single data frame
```{r}
month_rides <- rbind(
  casual_riders %>% 
    mutate(month = as.factor(format(as.Date(started_at), "%m")), rider_type = "Casual Rider") %>% 
    group_by(month, rider_type) %>% 
    summarize(count = n()),
  annual_members %>% 
    mutate(month = as.factor(format(as.Date(started_at), "%m")), rider_type = "Annual Member") %>% 
    group_by(month, rider_type) %>% 
    summarize(count = n()) )
```

# Plot the bar chart in month
```{r}
ggplot(data = month_rides, aes(x = month, y = count, fill = rider_type)) +
  geom_col(position = "dodge") +
  labs(x = "Month", y = "Number of Rides", title = "Number of Rides by Month and Rider Type") +
  scale_fill_manual(values = c("red", "blue"), labels = c("Casual Riders", "Annual Members")) +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") +
  scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")) +
  scale_y_continuous(labels = scales::comma)
```

#-------------------------------------------------------
#   3.C - Member behavior vs time intervals of the day
#-------------------------------------------------------

# A bar chart to show the number of rides taken by both casual riders and annual members at different 3 hour time intervals:

# Create a new column in your data frame with 3 hour time intervals
```{r}
all_data$time_interval <- cut(all_data$started_at, breaks = "3 hours")
```

# Count the number of rides for each time interval and type of member
```{r}
ride_counts <- 
  all_data %>% 
  group_by(time_interval, member_casual) %>% 
  tally()
```

```{r}
summary(time_interval)
```

# Plot the bar chart
```{r}
ggplot(all_data, aes(x = format(started_at, "%H"), fill = member_casual)) +
  geom_bar(position = "dodge") +
  scale_x_discrete(limits = c("00", "03", "06", "09", "12", "15", "18", "21")) +
  xlab("Hour of the day (24h format)") +
  ylab("Number of rides") +
  ggtitle("Number of rides taken by casual riders and annual members at different 3 hour intervals") +
  theme_minimal() + 
  theme(legend.position = "top") + 
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("red", "blue"), labels = c("Casual", "Annual"))
```

#-------------------------------------------------------------------
#   3.D - Rides taken by each member (normalized) by rideable type
#-------------------------------------------------------------------

# Calculate the number of rides taken by each member type and rideable type
```{r}
ride_counts <- all_data %>% 
  filter(rideable_type != "docked_bike") %>%
  group_by(member_casual, rideable_type) %>% 
  tally()
```

# Calculate the total number of rides taken by each member type
```{r}
total_rides <- all_data %>% 
  filter(rideable_type != "docked_bike") %>%
  group_by(member_casual) %>% 
  summarize(total = n())
```

# Join the total rides and ride count data frames
```{r}
ride_counts <- inner_join(ride_counts, total_rides, by = "member_casual")
```

# Normalize the number of rides by dividing by the total trips on each member type
```{r}
ride_counts$normalized <- ride_counts$n / ride_counts$total
```

# Plot the bar chart
```{r}
ggplot(ride_counts, aes(x = rideable_type, fill = member_casual, y = normalized)) +
  geom_bar(stat = "identity", position = "dodge") +
  xlab("Rideable Type") +
  ylab("Normalized Number of Rides") +
  scale_y_continuous(labels = scales::percent_format()) +
  ggtitle("Normalized number of rides taken by casual riders and annual members on different types of bikes") +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values = c("red", "blue"),
                    breaks = c("casual", "member"),
                    labels = c("Casual Member", "Annual Member"))
```

